\documentclass[11pt]{article}

\usepackage{xcolor}
\usepackage{fullpage}
\usepackage[colorlinks, allcolors=blue]{hyperref}
\usepackage{listings}
\usepackage{parskip}
\usepackage{amssymb}

\lstdefinelanguage{diff}{
  morecomment=[f][\color{blue}]{@@},           % group identifier
  morecomment=[f][\color{red}]{-},             % deleted lines
  morecomment=[f][\color{green!50!black}]{+},  % added lines
  morecomment=[f][\color{magenta}]{---},       % diff header lines
  morecomment=[f][\color{magenta}]{+++},
}

\lstset{
  basicstyle=\footnotesize\ttfamily,
}

\newcommand{\emailaddress}{marc.mutz@kdab.com}
\newcommand{\email}{\href{mailto:\emailaddress}{\emailaddress}}

\newcommand{\wgpaper}[1]{\href{https://wg21.link/#1}{#1}}
\newcommand{\isref}[1]{\textbf{[\wgpaper{#1}]}}
\newcommand{\isnref}[2]{\textbf{[\href{https://wg21.link/#1\##2}{#1}]/#2}}

\date{}
\title{Allowing trailing commas in \emph{ctor-initializer}}

\begin{document}

\maketitle\vspace{-2cm}

\begin{tabular}{ll}
  Document \#:&P2067R0\\
  Date:       &\today\\
  Project:    &Programming Language C++\\
              &EWG Incubator\\
%              &Library Group\\
  Reply-to:   &\author{Marc Mutz} \textless\email\textgreater
\end{tabular}
\vspace{1cm}
\begin{abstract}
  C++11 inherited trailing commas in \emph{enum-specifier} from C99,
  and trailing commas in \emph{braced-init-list} from even earlier C
  versions.

  C++ adds a list that does not allow trailing commas:
  \emph{ctor-initalizer}.

  We propose to allow a trailing comma, such that the following becomes valid C++:

  \begin{lstlisting}[language=C++]
    Type() :
        m_one(~~~), // trailing-comma style
        m_two(~~~),
    {}
  \end{lstlisting}
\end{abstract}


%\tableofcontents

\section{Motivation and Scope}

The author is unaware of the reasons the C committee decided to allow
trailing commas in \emph{enum-specifier}, but all reasons likely apply
to \emph{ctor-initalizer}, too, just more so, as for
\emph{enum-specifier}, one can always invent an artificial end
enumerator to keep ``production values'' trailing-comma-seprarated:

\begin{lstlisting}[language=C++]
  enum E {
    One,
    Two,

    NumE
  };
\end{lstlisting}

This option does not exist for \emph{ctor-initalizer}.

In our experience, the users of trailing-comma lists aim to avoid
touching unrelated lines of code when they perform changes using a
version control system.

Consider the following example, where a new data member is being added
at the end:

\begin{lstlisting}[language=diff]
    Type() :
        m_one(~~~),
-       m_two(~~~)
+       m__two(~~~),
+       m_three(~~~)
    {}
\end{lstlisting}

The \texttt{m\_two} line is changed, but not functionally. This is
frowned upon when using VCSs, because it disturbs the history of
changes, which, in widely-used VCSs such as git, svn, etc., is
recorded line-by-line. In reports like \texttt{git blame}, the
\texttt{m\_two} initialisation will appear as if it has been changed
when \texttt{m\_three} was added, and a deeper inquiry (looking at the
exact commit) is needed to find that just a comma was added.

Same diff with this proposal accepted:

\begin{lstlisting}[language=diff]
    Type() :
        m_one(~~~),
        m_two(~~~),
+       m_three(~~~),
    {}
\end{lstlisting}

As a work-around, the following style has been adopted, at least in
parts of Qt, and probably elsewhere:

\begin{lstlisting}[language=C++]
    Type()
      : m_one(~~~)
      , m_two(~~~) // leading-comma style
    {}
\end{lstlisting}

This style, however, still treats one member special: If the first
member is removed, the diff still touches the unrelated line
initialising \texttt{m\_two}:

\begin{lstlisting}[language=diff]
    Type()
-     : m_one(~~~)
-     , m_two(~~~)
+     : m_two(~~~)
    {}
\end{lstlisting}

The reason why this style is still preferred, even though it doesn't
solve all of the problem, is that removal of the first data member is
statistically less likely than adding a new data member at the end.

The matter was previously discussed in 2015 on the
\texttt{std-proposals} mailing-list\cite{mld}. Arthur O'Dwyer
identified the changes required to the standard\cite{std}, and Ville
developed a patch implementing this for GCC\cite{GCC}, but it appears
that no proposal was put forth in the aftermath.

This aims to be that missing proposal.

While some participants of the the \texttt{std-proposals} 2015
discussion questioned whether the work involved in adapting all C++
parsers was worth the gain, the authors believe that the recurring
nature of this topic, as well as the avidity with which users attempt
to work around the issue, warrants a settlement by allowing trailing
commas.

\section{Impact on the Standard}

Minimal. The syntax we propose to make valid was ill-formed before.

\section{Proposed Wording}

\subsection{Changes to \cite{cpp2a}}

In \isnref{class.base.init}1, as well as in \isref{gram.class}, change
the production for \emph{ctor-initalizer} as indicated:

  \emph{
  ctor-initalizer:\\
  \hspace*{1cm}: mem-initializer-list\colorbox{green}{ ,$_{opt}$}
  }

\subsection{Feature Macro}

We propose to use a new macro,
\texttt{\_\_cpp\_ctor\_trailing\_commas}, to indicate an
implementation's support for this feature.

\section{Design Decisions}

\subsection{Extension to other lists}

One of the points that was raised in the \texttt{std-proposals} 2015
discussion was whether to allow trailing commas for all lists,
incl.\ e.g.\ function arguments. We welcome such discussion, but deem
it outside the scope of this proposal for at least its first
iteration, mainly because there is no general ``list'' grammar
production in the standard to which a change could be uniformly
applied, and other types of lists have not seen the same kinds effort
being spent on work-arounds that \emph{ctor-initalizer} has
``enjoyed''.

\section{Acknowledgements}

Arthur O'Dwyer provided the wording on which this proposal is based.\cite{std}

\section{References}
\renewcommand{\section}[2]{}%
\begin{thebibliography}{9}
\bibitem{mld}
  \url{https://groups.google.com/a/isocpp.org/d/msg/std-proposals/I8N_75J9ZB8/9Mm27qvbYZkJ}
\bibitem{std}
  Arthur O'Dwyer\newline
  Permit mem-initializer-list to end with a trailing comma\newline
  \url{https://github.com/cplusplus/draft/commit/5417003045bad50705847f3d20ff72d78aeec32a}
\bibitem{GCC}
  Ville Voutilainen\newline
  GCC Patch ``Allow trailing comma in a mem-initializer-list''\newline
  \url{https://github.com/villevoutilainen/gcc/commit/3e7dc8cc67ef5a3c302ff15c90b0c7cbc56760e9}
\bibitem[N4820]{cpp2a}
  Richard Smith (editor)\newline
  \emph{Working Draft: Standard for Programming Language C++}\newline
  \url{http://wg21.link/N4820}
\end{thebibliography}

\end{document}

