\documentclass[11pt]{article}

\usepackage{xcolor}
\usepackage{fullpage}
\usepackage[colorlinks, allcolors=blue]{hyperref}
\usepackage{listings}
\usepackage{parskip}

\lstdefinelanguage{diff}{
  morecomment=[f][\color{blue}]{@@},           % group identifier
  morecomment=[f][\color{red}]{-},             % deleted lines
  morecomment=[f][\color{green!50!black}]{+},  % added lines
  morecomment=[f][\color{magenta}]{---},       % diff header lines
  morecomment=[f][\color{magenta}]{+++},
}

\lstset{
  basicstyle=\footnotesize\ttfamily,
}

\newcommand{\emailaddress}{marc.mutz@kdab.com}
\newcommand{\email}{\href{mailto:\emailaddress}{\emailaddress}}

\date{}
\title{Deduction Guides for packaged\_task}

\begin{document}

\maketitle\vspace{-2cm}

\begin{flushright}
  \begin{tabular}{ll}
  Document \#:&P1038R0\\
  Date:       &\date{2018-04-19}\\
  Project:    &Programming Language C++\\
              &Library Evolution Group\\
%              &Library Group\\
  Reply-to:   &\author{Marc Mutz}\\
              &\textless\email\textgreater
  \end{tabular}
\end{flushright}

\section{Introduction}

\texttt{std::function} has deduction guides, but
\texttt{std::packaged\_task}, which is otherwise very similar to
\texttt{std::function}, does not. We propose to add deduction guides
for \texttt{packaged\_task} that are compatible with the ones for
\texttt{function}.

\section{Impact on the Standard}

Minimal. We propose to add deduction guides to
\texttt{packaged\_task}, which lacked them. This enables code that did
not previously compile, but does not change the meaning of existing code.

\section{Proposed Wording}

NB: The intent is to re-use the wording that \texttt{std::function} has in
the draft at the time of acceptance.

In \texttt{[futures.task]}, before the non-member
\texttt{swap} function declaration, add:

\begin{quote}
\begin{lstlisting}[language=C++]
  template<class R, class... ArgTypes>
    packaged_task(R(*)(ArgTypes...)) -> packaged_task<R(ArgTypes...)>;

  template<class F> packaged_task(F) -> packaged_task<see below>;
\end{lstlisting}
\end{quote}

In \texttt{[futures.task.members]}, after the
\texttt{packaged\_task(F\&\&)} constructor description, add:

\begin{quote}
\texttt{template<class F> packaged\_task(F) -> packaged\_task<see below>;}

\textit{Remarks:} This deduction guide participates in overload
resolution only if \texttt{\&F​::​operator()} is well-formed when treated
as an unevaluated operand.  In that case, if
\texttt{decltype(\&F​::​operator())} is of the form
\texttt{R(G​::​*)(A...) cv \&$_{opt}$ noexcept$_{opt}$} for a class type
\texttt{G}, then the deduced type is \texttt{packaged\_task<R(A...)>}.
\end{quote}

\section{Acknowledgements}

Barry Revzin found a bug in a draft version of this paper. All other
errors are mine.

%\section{References}
%\renewcommand{\section}[2]{}%
%\begin{thebibliography}{9}
%
%  \bibitem[N3797]{N3797}
%    Stefanus Du Toit,
%    \emph{Working Draft, Standard for Programming Language C++}\newline
%    \url{http://www.open-std.org/jtc1/sc22/wg21/docs/papers/2013/n3797.pdf}
%
%\end{thebibliography}

\end{document}

